Utilizing Google's Mobility Report to understand Infectious Disease Dynamics

### 2025-02-21 code/eda-code/eda.qmd
Project Part 2
Might be a bit bare, however, I took caution to fleshing out too much if my topic needs to be adjusted. Right now, I have considered adding in the CDC Vaccination Hesitancy survey (possibly to map out and compare with mobility dynamics) or the County Health Ranking (possibly to explore how communities of differing SES dealt with cases in terms of mobility).

---
title: "An example exploratory analysis script"
date: "2024-02-07"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
```

Load the data.

```{r}
# importing data
mobility_data_2020 <- read.csv(here("data", "raw-data", "2020_US_Region_Mobility_Report.csv"))
mobility_data_2021 <- read.csv(here("data", "raw-data", "2021_US_Region_Mobility_Report.csv"))
mobility_data_2022 <- read.csv(here("data", "raw-data", "2022_US_Region_Mobility_Report.csv"))
combined_mobility_data <- bind_rows(mobility_data_2020, mobility_data_2021, mobility_data_2022)

time_series_covid <- read.csv(here("data", "raw-data", "time_series_covid19_confirmed_US.csv"))

chr_data_2024 <- read.csv(here("data", "raw-data", "chr_trends_csv_2024.csv"))

@@

# Data exploration through graphs

Showing a bit of code to produce and save a summary table.

```{r}
# graph covid cases overtime

# Select only the date columns (starting from the 12th column)
date_columns_cases <- names(time_series_covid)[12:ncol(time_series_covid)]

# Ensure column names are correctly formatted as dates (removing 'X' from the date string)
date_fixed_cases <- as.Date(gsub("X", "", date_columns_cases), format = "%m.%d.%y")  # Fixed missing parenthesis

# Sum case counts across all locations for each date
total_cases <- colSums(time_series_covid[, date_columns_cases, drop = FALSE], na.rm = TRUE)

# Convert to a tidy data frame
total_cases_df <- data.frame(Date = date_fixed_cases, Cases = total_cases)

# Plot the total case counts over time with formatted y-axis labels without using 'scales'
case_count_graph <- ggplot(total_cases_df, aes(x = Date, y = Cases)) +
  geom_line(color = "blue") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()

print(case_count_graph)
```

```{r}
# Calculate the daily new cases by subtracting previous day's total from the current day's total
new_cases <- c(0, diff(total_cases))  # diff() gives the difference between consecutive days; first day is set to 0

# Convert to a tidy data frame for new cases
new_cases_df <- data.frame(Date = date_fixed_cases, New_Cases = new_cases)

# Plot the new cases (incidence) over time with formatted y-axis labels
incidence_graph <- ggplot(new_cases_df, aes(x = Date, y = New_Cases)) +
  geom_line(color = "red") + 
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  # Format y-axis labels
  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "New Cases per Day") +
  theme_minimal()

print(incidence_graph)
```

```{r}
# Filter for U.S. data (adjust based on actual column name for region/country)
us_mobility_data <- combined_mobility_data %>%
  filter(country_region == "United States", sub_region_1 == "")

# Ensure the Date column is in Date format
us_mobility_data$Date <- as.Date(us_mobility_data$date, format = "%Y-%m-%d")  # Adjust if needed

# Calculate the 7-day moving average for each mobility type
us_mobility_data <- us_mobility_data %>%
  mutate(
    grocery_and_pharmacy_avg = rollmean(grocery_and_pharmacy_percent_change_from_baseline, 7, fill = NA, align = "right"),
    transit_stations_avg = rollmean(transit_stations_percent_change_from_baseline, 7, fill = NA, align = "right"),
    workplaces_avg = rollmean(workplaces_percent_change_from_baseline, 7, fill = NA, align = "right"),
    residential_avg = rollmean(residential_percent_change_from_baseline, 7, fill = NA, align = "right")
  )

# Plot the 7-day moving averages for each mobility type
mobility_graph <- ggplot(us_mobility_data, aes(x = Date)) +
  geom_line(aes(y = grocery_and_pharmacy_avg, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = transit_stations_avg, color = "Transit Stations")) +
  geom_line(aes(y = workplaces_avg, color = "Workplaces")) +
  geom_line(aes(y = residential_avg, color = "Residential")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
    x = "Date",
    y = "7-Day Moving Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_minimal()

print(mobility_graph)
```
```{r}

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

Histogram plots for the continuous outcomes.

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results", "figures", "height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

>>>>

---
title: "Exploratory Data Analysis"
date: "2025-02-21"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.
# Setup

```{r}
#load needed packages. make sure they are installed
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
library(tigris)
library(sf)
```

Load the data from processingfile-v1 file.

```{r}

# processed data
processeddata <- readRDS(here("data", "processed-data", "processeddata.rds"))

@@

Creation of a heatmap showing total COVID 19 cases on the last observed date, 10/15/2022. I also created a heat map showing mobility changes, with hotter indicating more mobility to retail and recreational spots. Interestingly, the two maps seem to correlate even though I used total COVID-19 cases rather than new ones.
```{r}
# Load the state-level shapefile 
us_map_states <- tigris::states(cb = TRUE) %>%
  st_as_sf()  # Convert the shapefile to a Simple Feature object

# Clean the state names in the shapefile
us_map_clean <- us_map_states %>%
  mutate(NAME = tolower(NAME))  # Convert state abbreviation to lowercase

# Filter your dataset to get data for the last recorded day and group by state
last_day <- max(processeddata$date)
data_last_day <- processeddata %>%
  filter(date == last_day) %>%
  group_by(state) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE),  # Sum of total cases
            retail_and_recreation_percent_change_from_baseline = 
              mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE)) %>%  # Average mobiltiy changes
  mutate(state = tolower(state))  # Convert state names to lowercase

# Join the shapefile with the COVID-19 data
us_map_data <- us_map_clean %>%
  left_join(data_last_day, by = c("NAME" = "state"))

# plot total COVID-19 cases to USA map on 10/15/2022
cases_map <- ggplot(us_map_data) +
  geom_sf(aes(fill = total_cases), color = "white") +
  scale_fill_viridis_c(option = "plasma", trans = "log", labels = NULL) + 
  labs(title = paste("Heatmap of Total COVID-19 Cases in the U.S. on", last_day),
       fill = "Hotter = More Cases") +
  theme_minimal() +
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))

print(cases_map)

figure_file = here("results","figures","casemap.png")
ggsave(filename = figure_file, plot=cases_map)

# Create the heatmap for retail and recreation mobility changes
mobility_map <- ggplot(us_map_data) +
  geom_sf(aes(fill = retail_and_recreation_percent_change_from_baseline), color = "white") +  
  scale_fill_viridis_c(option = "inferno", trans = "identity") +
  labs(title = paste("Heatmap of Retail and Recreation Mobility Changes in the U.S. on", last_day),
       fill = "Mobility Change") +
  theme_minimal() +
  theme(
        legend.spacing.y = unit(1, "cm"),  
        legend.box.spacing = unit(1, "cm"),  
        legend.key.size = unit(1, "cm")) +  
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))

print(mobility_graph)

# save graph
figure_file = here("results","figures","mobilitymap.png")
ggsave(filename = figure_file, plot=mobility_map)
```

This graph aims to plot total cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all total cases.
```{r}
# Create new data frame to have cases per day instead of state/county
total_cases_df <- processeddata %>%
  group_by(date) %>%
  summarise(Cases = sum(total_cases, na.rm = TRUE)) %>%
  ungroup()

# Plot the total case counts over time
case_count_graph <- ggplot(total_cases_df, aes(x = date, y = Cases)) +
  geom_line(color = "blue") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  
  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()

print(case_count_graph)

# save graph
figure_file = here("results","figures","casecountplot.png")
ggsave(filename = figure_file, plot=case_count_graph)
```
This graph aims to plot new cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all new cases.
```{r}
# data frame to sum all cases on a given day for every location
new_cases_df <- processeddata %>%
  group_by(date) %>%
  summarise(Cases = sum(new_cases, na.rm = TRUE)) %>%
  ungroup()

# Plot the new cases over time
incidence_graph <- ggplot(new_cases_df, aes(x = date, y = Cases)) +
  geom_line(color = "red") + 
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "New Cases per Day") +
  theme_minimal()

print(incidence_graph)

# save graph
figure_file = here("results","figures","incidenceplot.png")
ggsave(filename = figure_file, plot=incidence_graph)

```
I aim to plot out the changes in mobility for each category as time progresses. Originally, it plotted out everyday, however, the line was far too jagged so I applied a weekly rolling average instead. I also had to create a dataframe to average values for a given day for every location.
```{r}

# Create dataframe of average mobility for everywhere in the US per date
us_average_mobility <- processeddata %>%
  group_by(date) %>%
  summarise(
    total_mobility_changes = mean(average_mobility, na.rm = TRUE),
    transit_mean = mean(transit_stations_percent_change_from_baseline, na.rm = TRUE),
    parks_mean = mean(parks_percent_change_from_baseline, na.rm = TRUE),
    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(date) %>%  # Ensure dates are sorted
  mutate(
    transit_mean_7day = rollmean(transit_mean, k = 7, fill = NA, align = "right"),
    parks_mean_7day = rollmean(parks_mean, k = 7, fill = NA, align = "right"),
    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
  )

# Plot the 7-day averages for each mobility type
mobility_graph <- ggplot(us_average_mobility, aes(x = date)) +
  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = transit_mean_7day, color = "Transit Stations")) +
  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
  geom_line(aes(y = parks_mean_7day, color = "Recreational")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
    x = "Date",
    y = "7-Day Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_minimal()

print(mobility_graph)

# save graph
figure_file = here("results","figures","mobilitygraph.png")
ggsave(filename = figure_file, plot=mobility_graph)
```

### 2025-02-21 products/manuscript/Manuscript.qmd
Project Part 2
Might be a bit bare, however, I took caution to fleshing out too much if my topic needs to be adjusted. Right now, I have considered adding in the CDC Vaccination Hesitancy survey (possibly to map out and compare with mobility dynamics) or the County Health Ranking (possibly to explore how communities of differing SES dealt with cases in terms of mobility).

---
title: "Utilizing Google's Mobility Report to predict outbreak dynamics"
author: "Vincent Nguyen"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

@@

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some\@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

@@

The County Health Ranking is a data set that provides a comprehensive overview of various health factors and outcomes across U.S. counties. Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more.\
\
\

@@

Investigate if counties with similar health rankings but different mobility trends (or vice versa) experienced different outbreak severities, potentially revealing the effectiveness of local interventions \[reword\]

Examine the time lag between changes in mobility and subsequent changes in infection rates \[reword\]

@@

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

@@

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

@@

```{r}

@@

Possibly utilize the EpiModeling package to understand infectious disease dynamics based on the analysis and then simulate future outbreaks? I am not actually sure if this is a feasible use of the data.

@@

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

@@

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

@@

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

>>>>

---
title: "Utilizing Google's Mobility Report to understand Infectious Disease Dynamics"
author: "Vincent Nguyen"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

@@

Some text was left from the template in order to adjust to any feedback given. The manuscript will be more fleshed out once the final project idea is approved.

@@

$\land$ Corresponding author: vln27447\@uga.edu

@@

The County Health Ranking is a data set that provides a comprehensive overview of various health factors and outcomes across U.S. counties. Data collected includes metrics on healthcare access, insurance coverage, income levels, violent crime, air and water quality, transportation access, and more. \[Not sure about the inclusion of this\]

The CDC Vaccination hesitancy survey is a possible data set I could use to make the analysis more complex.\
\
\

@@

Investigate if counties with similar health rankings but different mobility trends experienced different outbreak severities, potentially revealing the effectiveness of local interventions

Examine ifthere was lag time between changes in mobility and subsequent changes in infection rates

@@

@@

@@

```{r}
library(here)

@@

@@

To gain a clearer perspective on the overall scale of the pandemic, we visualized the total number of COVID-19 cases over time. This plot provides a overview of how case counts evolved.The graph reveals a large spike in cases in the beginning of 2022. This does correspond with figures further.

```{r}
#| label: casecountplot
#| fig-cap: "Total COVID-19 Cases overtime"
#| echo: FALSE
knitr::include_graphics(here("results","figures","casecountplot.png"))
```

To gain a clearer perspective on the overall scale of the pandemic, we visualized new caes of COVID-19 over time. This plot provides a overview of how case counts and infection evolved.The graph reveals a large spike in cases in the beginning of 2022 which corresponds with the total case count graph.

```{r}
#| label: incidenceplot
#| fig-cap: "Daily New COVID-19 Cases overtime in the U.S."
#| echo: FALSE
knitr::include_graphics(here("results","figures","incidenceplot.png"))
```

To gain a deeper understanding of how mobiltiy patterns shifted throughout the pandemic, weekly changes in mobility were visualized. This plot highlights fluctuations in movement across various sectors. Mobility was increased towards residential and recreational areas which warrants further exploration. This can help offer insight into how behavior towards COVID-19 restriction interventions evolved over time.

```{r}
#| label: mobiltiygraph
#| fig-cap: "Percent Changes in Mobility per week in the U.S."
#| echo: FALSE
knitr::include_graphics(here("results","figures","mobilitygraph.png"))
```

As hypothesized, an rise in COVID-19 case count is expected to lead to a decline in mobility to non-residential locations, as individuals alter their behavior in resposne to increasing risks. To explore this idea, total case count and changes in mobility were visualzied across the United States.\
\

```{r}
#| label: casemap
#| fig-cap: "Total COVID-19 Cases on 10/15/2022 mapped to the U.S."
#| echo: FALSE
knitr::include_graphics(here("results","figures","casemap.png"))
```

```{r}
#| label: mobilitymap
#| fig-cap: "Average of mobility changes on 10/15/2022 mapped to the U.S."
#| echo: FALSE
knitr::include_graphics(here("results","figures","mobilitymap.png"))
```

Analysis of the mapped data reveals a trend: states with higher case counts generally experienced more significant decreases in mobility. This suggests a correlation between rising COVID-19 case count and a greater reduction in movement to non-residential locations.

@@

@@
